A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution \thanks{A preliminary version of this work has appeared in the Proceedings of the 26th International Symposium on Graph Drawing and Network Visualization (GD 2018). This work is supported by the DFG grant Ka812/17-1}

### 2018-09-14 paper/cr-res-north.tex
Work on the journal version.

\documentclass{comjnl}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,pgfplots}
\usepackage[font=small]{subfig,caption}
%\usepackage{algorithm2e}
\usepackage{todonotes,lineno,paralist,soul}
\usepackage{rotating,hyperref,multicol}
\usepackage{booktabs,tabularx}
\graphicspath{{figures/}}

% ==================================================================
% Format tiks and pgfplots
\tikzstyle{densely dashed}=[dash pattern=on 0.75pt off 0.5pt]
\pgfplotsset{ytick style={draw=none}}
\renewcommand{\axisdefaultheight}{185pt}
\pgfkeys{/pgf/number format/.cd,1000 sep={}}
\pgfplotscreateplotcyclelist{mylist}{%
{black,thick,smooth},
{black!50!white,thick,smooth},
{black,thick,densely dashed,smooth},
{black!50!white,thick,densely dashed,smooth},
{black,double,smooth}
}
\pgfplotsset{
        table/search path={data},
    }
\pgfplotsset{
  /pgfplots/xlabel near ticks/.style={
     /pgfplots/every axis x label/.style={
        at={(ticklabel cs:0.5)},anchor=near ticklabel
     }
  },
  /pgfplots/ylabel near ticks/.style={
     /pgfplots/every axis y label/.style={
        at={(ticklabel cs:0.5)},rotate=90,anchor=near ticklabel}
     }
}
% ==================================================================
\newcommand{\myparagraph}[1]{\smallskip\noindent\textbf{#1}.}
% ==================================================================

\begin{document}

\author{Michael~A.~Bekos} 
\author{Henry~F\"orster}
\author{Christian~Geckeler} 
\author{Lukas~Holl\"ander} 
\author{Michael~Kaufmann}
\author{Amad\"aus~M.~Spallek} 
\author{Jan~Splett}

\title[Drawings of Graphs with High Crossing Resolution]{A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution}

\affiliation{
Institut f\"ur Informatik, Universit\"at T\"ubingen, T\"ubingen, Germany\\
}

\email{\{bekos,foersth,mk\}@informatik.uni-tuebingen.de\\\{geckeler,jan-lukas.hollaender,amadaeus.spallek,jan.splett\}@student.uni-tuebingen.de}

\received{13 September 2018}

\shortauthors{M.~A.~Bekos et al.}

\keywords{Graph Drawing, Angular, Crossing and Total Resolution.}

%\copyrightyear{2018} \vol{00} \issue{0} \DOI{000}
% ==================================================================

\begin{abstract}
The \emph{crossing resolution} of a non-planar drawing of a graph is the value of the minimum angle formed by any pair of crossing edges. Recent experiments suggest that the larger the crossing resolution is, the easier it is to read and interpret a drawing of a graph. However, maximizing the crossing resolution turns out to be an NP-hard problem in general, and only heuristic algorithms are known that are mainly based on appropriately adjusting force-directed algorithms.

In this paper, we propose a new heuristic algorithm for the crossing resolution maximization problem and we experimentally compare it against the known approaches from the literature. Our experimental evaluation indicates that the new heuristic produces drawings with better crossing resolution, but this comes at the cost of slightly higher aspect ratio, especially when the input graph is large. 
\end{abstract}

% ==================================================================
\maketitle
% ==================================================================

% ==================================================================
\section{Introduction}
\label{sec:introduction}
% ==================================================================

In Graph Drawing, there exists a rich literature and a wide range of techniques for drawing planar graphs; see, e.g.,~\cite{DBLP:journals/combinatorica/FraysseixPP90,DBLP:conf/gd/GutwengerM98,DBLP:journals/algorithmica/Kant96}. However, drawing a non-planar graph, and in particular when it does not have some special structure (e.g., degree restriction), is a quite difficult and very challenging task, mainly due to the edge crossings that negatively affect the drawing's quality~\cite{DBLP:journals/iwc/Purchase00}. As a result, the established techniques are significantly fewer (e.g., crossing minimization heuristics~\cite{DBLP:journals/algorithmica/EadesW94,DBLP:journals/tsmc/SugiyamaTT81}, energy-based layout algorithms~\cite{DBLP:journals/congnum/Eades84,DBLP:journals/spe/FruchtermanR91}); for an overview we point the reader to~\cite{DBLP:books/ph/BattistaETT99,DBLP:conf/dagstuhl/1999dg,DBLP:reference/crc/2013gd}.

In this context, Huang et al.~\cite{DBLP:conf/apvis/Huang07,DBLP:journals/vlc/HuangEH14} a decade ago introduced some important experimental evidence (through eye-tracking experiments), that edge crossings may not negatively affect the drawing's quality too much (and hence the human's ability to read and interpret it), when the angles formed by the crossing edges are large. In other words, while prior to these experiments it was commonly accepted that mainly the number of crossings is the most important parameter for judging the quality of a non-planar graph drawing~\cite{DBLP:conf/gd/Purchase97,DBLP:journals/iwc/Purchase00}, it turned out that the types of edge crossings also matter. As a result, a new and prominent research direction was initiated, recognized under the term ``beyond planarity''~\cite{Shonan2016,Dagstuhl2016,SoCG2017}, which focuses on graphs and their properties, when different constraints on the types of edges crossings are imposed; refer to~\cite{DBLP:journals/corr/abs-1804-07257} for a recent survey.

Formally, the value of the minimum angle formed by any two crossing edges in a drawing is referred to as its \emph{crossing resolution\footnote{Also, refered to as \emph{crossing angle resolution} in the literature.}}. Analogously, the crossing resolution of a graph is defined as the maximum crossing resolution over all its drawings. Clearly, the crossing resolution of a non-planar graph is at most $90^\circ$, while a graph that admits a drawing with crossing resolution $90^\circ$ is called \emph{right-angle-crossing} (or \emph{RAC}, for short) graph; see Figure~\ref{fig:examples} for an illustration. Notably, RAC graphs are sparse (an $n$-vertex RAC graph has at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}), while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}.

\begin{figure}[t!]
	\centering
	\subfloat[\label{fig:k5} {}]{
	\includegraphics[page=1]{figures/examples}}
	\hfil
	\subfloat[\label{fig:k6} {}]{
	\includegraphics[page=2]{figures/examples}}
	\caption{Illustration of: %
	(a)~a RAC drawing of the complete graph $K_5$, and
	(b)~a drawing of the complete graph $K_6$, whose crossing resolution is arbitrarily close to $90^\circ$.}
	\label{fig:examples}
\end{figure}

The latter result is an indication that the problem of finding drawings with high crossing resolution might also be difficult, even though, formally, its complexity has not been settled yet for values of the crossing resolution smaller than $90^\circ$. It is also worth noting that the literature is significantly more limited, when restricting the crossing resolution to be smaller than $90^\circ$, as also evidenced by Section~\ref{sec:relatedwork}.

From a practical point of view, we are only aware of two methods that aim at drawings with high crossing resolution; both of them are adjustments of force-directed algorithms~\cite{DBLP:journals/congnum/Eades84}. The first one is due to Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, while the~second one is due to Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. Common in both algorithms is that they apply appropriate forces on the endvertices of every pair of crossing edges. Each of them uses a different way to compute (the direction and the magnitude of) the forces, but the underlying idea of both is the same: the smaller the crossing angles are, the larger are the magnitudes of the forces applied at their endvertices.

In this work, we approach the crossing resolution maximization problem from a different perspective. We suggest a simple and intuitive randomization method for computing drawings with high crossing resolution, which, in a sense, mimics the way a human would try to increase the crossing resolution of a drawing. How would one increase the crossing resolution of a given drawing? First, she would try to identify the pair of edges that define the crossing resolution of the drawing (we call them \emph{critical} edges); then, she would try to move an endvertex of this pair (which we choose at random), hoping that by this move the crossing resolution will increase. Of course, we cannot consider all possible positions for the vertex to be moved. Instead, we consider a small set of randomly generated ones. If there exists a position among them, that does not lead to a reduction of the crossing resolution, we move the vertex to this position.

In general, randomization is a technique that has not been deeply examined in Graph Drawing, as it seems difficult to even speculate about the expected quality of the produced drawings; a notable exception is the randomized approach by Goldschmidt and Takvorian~\cite{DBLP:journals/networks/GoldschmidtT94} for computing large planar subgraphs. Since we also could not provide any theoretical guarantee on the expected quality of the produced drawings (mainly due to the nature of the problem itself),
we followed a more practical approach. We implemented our algorithm and the force-directed ones of~\cite{DBLP:journals/cj/ArgyriouBS13} and~\cite{DBLP:journals/vlc/HuangEHL13}, and we experimentally compared them on standard benchmark graphs that are widely used in Graph Drawing for comparing different drawing algorithms. Our evaluation indicates that our method significantly outperforms the force-directed algorithms~\cite{DBLP:journals/cj/ArgyriouBS13,DBLP:journals/vlc/HuangEHL13} in terms of crossing resolution, but this comes at the cost of slightly worse running time, especially for large and dense graphs. Analogous results are obtained, when our algorithm and the ones of~\cite{DBLP:journals/cj/ArgyriouBS13} and~\cite{DBLP:journals/vlc/HuangEHL13} are adjusted to maximize the \emph{angular resolution} (that is, the minimum value of the angle between any two adjacent edges~\cite{DBLP:journals/siamcomp/FormannHHKLSWW93}) or the \emph{total resolution} (that is, the minimum of the angular and the crossing resolution~\cite{DBLP:journals/cj/ArgyriouBS13}).

\medskip\noindent{\it Preliminaries:}
Unless otherwise specified, in this paper we consider simple undirected graphs. Let $G=(V,E)$ be such a graph. The degree of vertex $u\in V$ of $G$ is denoted by $d(u)$. The degree $d(G)$ of  graph $G$ is defined as the maximum degree of its vertices, i.e., $d(G)=\max_{u\in V}d(u)$. Given a drawing $\Gamma(G)$ of $G$, we denote by $p(u)=(x_u,y_u)$ the position of vertex $u \in V$ of $G$ in $\Gamma(G)$. %For a pair of vertices $u,v\in V$ of $G$, the unit-length vector from $p(u)$ to $p(v)$ is denoted as $\overrightarrow{p(u)p(v)}$, while the line segment delimited by $p(u)$ and $p(v)$ as $\overline{p(u)p(v)}$.

\medskip\noindent{\it Structure of the paper:}
The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works. Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project.

% ==================================================================
\section{Related Work}
\label{sec:relatedwork}
% ==================================================================

As already mentioned, the study of the crossing resolution maximization problem has mainly focused on its optimal case, i.e., on the study of RAC graphs. An $n$-vertex RAC graph has at most $4n-10$ edges~\cite{DBLP:journals/tcs/DidimoEL11}, while deciding whether a graph is RAC is NP-hard~\cite{DBLP:journals/jgaa/ArgyriouBS12}. The maximally-dense RAC graphs are 1-planar~\cite{DBLP:journals/dam/EadesL13}, i.e., they can be drawn with at most one crossing per edge. Actually, several~relationships between the class of RAC graphs and subclasses of 1-planar graphs are known~\cite{DBLP:journals/dam/BachmaierBHNR17,DBLP:journals/tcs/BrandenburgDEKL16}. Deciding, however, whether a 1-planar graph is RAC is NP-hard~\cite{DBLP:journals/tcs/BekosDLMM17}. Note that the problem of finding RAC drawings has also been studied in the presence of bends~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/comgeo/ArikushiFKMT12,DBLP:journals/tcs/DidimoEL11,DBLP:journals/mst/GiacomoDLM11} and by imposing restrictions on the degree~\cite{DBLP:conf/s-egc/AngeliniBDFHKLL11}, the structure~\cite{DBLP:journals/ipl/DidimoEL10} and the drawing~\cite{DBLP:journals/algorithmica/GiacomoDEL14,DBLP:conf/wg/HongN15} of the graph. The results are fewer, when the right-angle constraint is relaxed. Dujmovic et al.~\cite{DBLP:journals/cjtcs/DujmovicGMW11} proved that an $n$-vertex graph with crossing resolution at least $\alpha$ radians, has at most $(3n-6)\pi/\alpha$ edges. Corresponding density results are also known in the presence of bends~\cite{DBLP:journals/siamdm/AckermanFT12,DBLP:journals/mst/GiacomoDLM11}.

An immediate observation emerging from the above overview is that the~focus has been primarily on theoretical aspects of the problem. Most of the approaches that could be useful in practice are based on force-directed techniques~\cite{DBLP:books/ph/BattistaETT99,DBLP:journals/congnum/Eades84}. COWA is a system that supports conceptual web site traffic analysis~\cite{DBLP:conf/apvis/DidimoLR10}; its algorithmic core is a force-directed heuristic to compute simultaneous embeddings of two non-planar graphs with high crossing resolution.
%
Didimo et al.~\cite{DBLP:conf/gd/DidimoLR10} describe topology-driven force-directed heuristics to achieve good trade-offs in terms of number of edge crossings, crossing resolution, and geodesic edge tendency; the obtained drawings, however, are not straight-line.
%
For straight-line drawings, Nguyen et al.~\cite{DBLP:conf/gd/NguyenEHH10} suggest a quadratic-program to increase the crossing angles of circular drawings.
%
Of more general scope are the already mentioned force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}.

%As already mentioned, the study of the crossing resolution maximization problem has mainly focused on (properties of) RAC graphs, that is, on the optimal case of the crossing resolution maximization problem. Their study was initiated by Didimo et al.~\cite{DBLP:journals/tcs/DidimoEL11}, who showed that an $n$-vertex RAC graph has at most $4n-10$ edges. Angelini et al.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11} presented results for two interesting variants, in which the input graph is either an acyclic digraph or a graph of bounded vertex degree. Argyriou et al.~\cite{DBLP:journals/jgaa/ArgyriouBS12} showed that the problem of deciding whether a graph is RAC is NP-hard. Angelini et al.~\cite{DBLP:conf/s-egc/AngeliniBDFHKLL11} observed that planar graphs of bounded degree admit straight-line RAC drawings in subquadratic area. Didimo et al.~\cite{DBLP:journals/ipl/DidimoEL10} characterized the complete bipartite RAC graphs. Di~Giacomo et al.~\cite{DBLP:journals/algorithmica/GiacomoDEL14}, and Hong and Nagamochi~\cite{DBLP:conf/wg/HongN15} studied variants, in which the vertices of the input graph are restricted on two parallel lines and on the circumference of a circle, respectively. Eades and Liotta~\cite{DBLP:journals/dam/EadesL13} showed that the optimal RAC graphs (i.e., those of maximum density) are 1-planar (i.e., they admit drawings in which each edge is crossed at most once). Deciding, however, whether a 1-planar graph is RAC is NP-hard~\cite{DBLP:journals/tcs/BekosDLMM17}. Other relationships between the class of RAC graphs and subclasses of 1-planar graphs have also been studied by Bachmaier et al.~\cite{DBLP:journals/dam/BachmaierBHNR17} and Brandenburg et al.~\cite{DBLP:journals/tcs/BrandenburgDEKL16}. Note that the problem of finding RAC drawings has been also studied in the presence of bends; see e.g.~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/comgeo/ArikushiFKMT12,DBLP:journals/tcs/DidimoEL11,DBLP:journals/mst/GiacomoDLM11}.

%To the best of our knowledge, there is only one work, by Dujmovic et al.~\cite{DBLP:journals/cjtcs/DujmovicGMW11}, which studies the crossing resolution maximization problem by relaxing the right-angle constraint on the crossing angles of the computed drawings. More precisely, in their work, Dujmovic et al.~\cite{DBLP:journals/cjtcs/DujmovicGMW11} proved that an $n$-vertex graph, whose crossing resolution is at least $\alpha$ (in radians) has at most $(3n-6)\pi/\alpha$ edges. Corresponding density results for the case, in which few bends are allowed along each edge, are known by Ackerman et al.~\cite{DBLP:journals/siamdm/AckermanFT12} and Di Giacomo et al.~\cite{DBLP:journals/mst/GiacomoDLM11}.

%An immediate observation emerging from the above literature overview is that the focus of the study of the crossing resolution maximization problem has been primarily on theoretical aspects of the problem. Most of the approaches that could be useful in practise are based on adjustments of force-directed techniques~\cite{DBLP:journals/congnum/Eades84}, according to which a graph is modelled as a physical system with forces acting on it, and a (good) drawing is obtained by an equilibrium of the system; for an introduction and a discussion of several variants refer to~\cite{DBLP:books/ph/BattistaETT99}. %Actually, the absence of approaches for computing drawings with high crossing resolution in practice is rather obvious and it can be justified, up to a certain point, by the fact that the problem turns to be NP-hard, when asking for drawings with optimal crossing resolution.
%
%More concretely, from an applicative point of view, Didimo et al.~\cite{DBLP:conf/apvis/DidimoLR10} describe a system, called \emph{COWA}, to support conceptual web site traffic analysis, whose algorithmic core is a force-directed heuristic to compute simultaneous embeddings of two non-planar graphs with high crossing resolution.
%
%In a follow up work, Didimo et al.~\cite{DBLP:conf/gd/DidimoLR10} describe heuristics, designed within the topology-driven force-directed framework, to achieve good trade-offs in terms of number of edge crossings, crossing resolution, and geodesic edge tendency.
%
%However, the obtained drawings are not straight-line. For straight-line drawings, Nguyen et al.~\cite{DBLP:conf/gd/NguyenEHH10} suggest a quadratic-programming based approach to increase the crossing angles of circular drawings.
%
%Of more general scope are the already mentioned works of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.\, which are also based on the force-directed technique.

% ==================================================================
\section{Description of our Heuristic Approach}
\label{sec:algorithm}
% ==================================================================

In this section, we describe our heuristic for obtaining drawings with high crossing resolution. The input of our heuristic consists of a graph $G$ and an initial drawing $\Gamma_0$ of $G$ with crossing resolution $c(\Gamma_0)$. We assume that no two edges of~$G$ overlap in $\Gamma_0$, i.e., $c(\Gamma_0)>0$. A circular drawing or a drawing obtained by applying a force-directed algorithm on $G$ clearly meets this precondition.

Our algorithm is iterative and at each iteration performs some operations that are mainly based on randomization. At the $i$-th iteration, we assume that we have computed a drawing $\Gamma_{i-1}$ of crossing resolution $c(\Gamma_{i-1}) \geq c(\Gamma_0)$.
%, where $\Gamma_0$ is our initial drawing.
In other words, we assume, as an invariant for our algorithm, that the crossing resolution cannot be decreased at some iteration. Then, a vertex of $\Gamma_{i-1}$ is chosen arbitrarily at random based on the so-called \emph{vertex-pool}, which may contain:
%
%\begin{itemize}
\begin{inparaenum}[(i)]
\item either all vertices of $\Gamma_{i-1}$, or
\item a prespecified subset of the vertices of $\Gamma_{i-1}$, called \emph{critical}.
\end{inparaenum}

Intuitively, the critical vertices are the endpoints of the edges that define the crossing resolution of drawing $\Gamma_{i-1}$. To formally define them, we first need to introduce the notion of critical edge-pairs. A pair of edges $e$ and $e'$ is called \emph{critical} in $\Gamma_{i-1}$, if $e$ and $e'$ cross in $\Gamma_{i-1}$ and the minimum angle that is formed at their crossing point is equal to $c(\Gamma_{i-1})$. The set of critical vertices of $\Gamma_{i-1}$ is then defined by the four endvertices of each critical edge-pair.

The role of critical vertices is central in our algorithm\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a good solution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a probabilistic random selection procedure, so that the vertices at graph-distance~$i$ from the ones of the vertex-pool have higher probability for selection than the corresponding ones at distance $j$  in the graph, when $0 \leq i<j$. So, if the vertex-pool contains only critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, the vertex-pool contains all vertices and each vertex can be chosen with the same probability.

%With this procedure, what we quickly realized from our experimental evaluation, is that by appropriately changing the location of a critical vertex at each iteration of our algorithm, the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm.
What we quickly realized from our practical analysis, is that the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. However, by focusing only at the critical vertices, it is~highly possible that the algorithm will get trapped to some local maxima after a number of iterations. So, special care is needed to avoid these bottlenecks, especially when the input graph is large. We will discuss ways to avoid them later in this section.

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Next, we describe how to compute its new position in the next drawing. Note that our method resembles probabilistic hill climbing approaches.

Let $v_i$ be the vertex of $\Gamma_{i-1}$ that has been chosen to be moved at the $i$-th iteration.
%Recall that the crossing resolution $c(\Gamma_{i})$ of drawing $\Gamma_{i}$ that is obtained after the $i$-th iteration of the algorithm must be at least as large as the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$ (by the invariant of our algorithm), that is, $c(\Gamma_i) \ge c(\Gamma_{i-1})$.
To compute the position of $v_i$ in the next drawing $\Gamma_i$, we consider a set of $\rho$ rays $r_0,r_1,\ldots,r_{\rho-1}$ that all emanate from $p(v_i)$ in $\Gamma_{i-1}$, such that the angle formed by ray $r_j$, with $j=0,1,\ldots,\rho-1$, and the horizontal axis equals to $2j\pi/\rho$, where $\rho>0$ is an integer parameter of the algorithm. These rays are then rotated by an angle that is chosen uniformly at random in the interval $[0,2\pi]$; see Fig.~\ref{fig:algo}. The position of vertex $v_i$ in $\Gamma_i$ will eventually be along one of the rays $r_0,r_1,\ldots,r_{\rho-1}$. More precisely, for each ray $r_i$ we choose a distance value $\delta_i$ uniformly at random from the interval $[\delta_{min},\delta_{max}]$, where $\delta_{min}$ and $\delta_{max}$ are two positive parameters of the algorithm. For each $j=0,1,\ldots,\rho-1$, a new point $\pi_j$ is obtained by translating $p(u)$ along $r_j$ by a distance $\delta_j$; point $\pi_j$ is \emph{feasible}, if the crossing resolution of the drawing obtained by placing vertex $v_i$ at $\pi_j$ and by keeping all other vertices of $G$ in their positions in $\Gamma_{i-1}$ is at least as large as the crossing resolution of $\Gamma_{i-1}$, and there is no vertex of $\Gamma_{i-1}$ at $\pi_j$.

\begin{figure}[t!]
	\centering
	\subfloat[\label{fig:algo-rays} {}]{
	\includegraphics[page=1, scale=0.9]{figures/algorithm}}
	\hfil
	\subfloat[\label{fig:algo-move} {}]{
	\includegraphics[page=2, scale=0.9]{figures/algorithm}}
	\caption{%
	Illustration of an iteration step of our algorithm:
	(a)~The chosen vertex is the white one;
	the computed rays $r_0,\ldots,r_7$ have been rotated by $8^\circ$;
	the black-colored points along these rays are points $\pi_0,\ldots,\pi_7$;
	among them, $\pi_4$ yields the best solution.
	(b)~The resulting drawing after moving the vertex at position $\pi_2$.}
	\label{fig:algo}
\end{figure}

If none of the points $\pi_j$, with $j=0,1,\ldots,\rho-1$ is feasible, then the position of $v_i$ in $\Gamma_i$ is $p(v_i)$, i.e., same as in $\Gamma_{i-1}$, since $c(\Gamma_i) \geq c(\Gamma_{i-1})$ must hold. If there is one or more feasible points, then one may consider two different approaches to determine the position of $v_i$ in $\Gamma_i$. The most natural is to choose the feasible point that maximizes the crossing resolution of the obtained drawing. As an alternative, one may rely again on randomization and chose uniformly at random one of the feasible points as the position of $v_i$ in $\Gamma_i$. We note that we did not observe any significant difference between these two approaches (in terms of the crossing resolution of the obtained drawings), so we simply adopted the first one. 
The termination condition of our algorithm is simple and depends on an in\-put parameter~$\tau$. More specifically, if the crossing resolution has not improved~during the last $\tau$ iterations, we assume that the algorithm has converged and we~stop.

% ==================================================================
\myparagraph{Avoiding local maxima}
% ==================================================================
%
To avoid getting trapped to locally optimal solutions, we mainly investigated two approaches, which are both  parametrizable by two input parameters $\zeta$ and $\zeta'$. The first mimics the human behaviour. What would~one do to escape from a locally optimal solution? She would stop trying to move the endvertices of the edges defining the crossing resolution; she would rather start moving ``irrelevant'' vertices hoping that
%by changing the embedding of the graph or its crossing structure,
by doing so a better solution will be easier to be computed afterwards. Our algorithm is mimicking this idea as follows:
%
\begin{inparaenum}[(i)]
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices.
\item if during the last $\zeta$ iterations the crossing resolution has not been~improved, then the vertex-pool becomes \emph{wider} by including all the vertices, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations;
\item afterwards, the vertex-pool switches back to the critical vertices.
\end{inparaenum}
%
%From our extensive practical analysis, we quickly realized that, while
While this approach turned out to be effective for smaller graphs, for graphs with more than 100 vertices, it was not so efficient; in most iterations with the wider vertex-pool, the embedding could not change in a beneficial way for the algorithm~to~proceed.

%What we observed is that in larger graphs it was unlikely to change the embedding of the graph in a beneficial way for the algorithm to proceed, mainly because of the random choice of the vertices to be moved; in most iterations with the wider vertex-pool, the vertices chosen to be moved were not in the part of the drawing where the critical vertices resided.

Our second approach is based on parameters $\rho$, $\delta_{min}$ and $\delta_{max}$ of the algorithm. Our idea was that if the algorithm gets trapped to a locally optimal solution, then a ``drastic'' or ``sharp'' move may help to escape. We turned this idea into an algorithmic implementation as follows:
%
\begin{inparaenum}[(i)]
\item if during the last $\zeta$ iterations the crossing resolution has not been improved, we double the values of $\rho$, $\delta_{min}$ and $\delta_{max}$, and the algorithm is executed with these values for $\zeta'$ iterations;
\item afterwards, $\rho$, $\delta_{min}$ and $\delta_{max}$ switch back to this initial value.
\end{inparaenum}
%
This approach may lead to drawings with larger area, but this is ``expected'', as it turns out that drawings with high crossing resolution may require large area~\cite{DBLP:journals/jgaa/AngeliniCDFBKS11,DBLP:journals/tcs/BrandenburgDEKL16}.

% ==================================================================
\myparagraph{Complexity issues}
% ==================================================================
%
A factor that highly affects the efficiency of our algorithm is the computation of the crossing points of the edges and the corresponding angles at these points. Given  a drawing, a na\"ive approach to compute its crossings requires $O(m^2)$ time, which can be improved by a plane-sweep technique to $O(m \log m + c)$ time, where $m$ and $c$ denote the number of edges and crossings.
%; see, e.g.,~\cite{DBLP:books/lib/BergCKO08}.

Instead of computing all crossing points and the corresponding angles for each candidate position of each iteration, we adopted a different approach for determining the set of feasible candidate positions, which turned out to be quite efficient in practice. Recall that we denoted by $v_i$ the vertex chosen at the $i$-th iteration step, and by $\pi_0,\ldots,\pi_{\rho-1}$ the candidate positions to move $v_i$. Let $e_0,\ldots,e_{d_i-1}$ be the edges incident to $v_i$, where $d_i=deg(v_i)$. Next, for each edge $e_k$ with $k=0,\ldots,d_i-1$  we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges in $\Gamma_{i-1}$. Let $\phi_i$ be the minimum crossing angle computed; this is our reference angle. Also, for each candidate position $\pi_j$ with $j=0,\ldots,\rho-1$, and for each edge $e_k$ with $k=0,\ldots,d_i-1$, we compute the crossings and the corresponding crossing angles of $e_k$ with all other edges of the drawing, assuming that $v_i$ is at $\pi_j$. Let $\chi_j$ be the minimum crossing angle computed with this approach, when $v_i$ is at position $\pi_j$. Clearly, $\pi_j$ is feasible only if $\chi_j \geq \phi_i$. Note that the complexity of this approach is $O(deg(v_i)m) = O(nm)$.

% ==================================================================
\subsection{Some Interesting Variants}
\label{ssec:variants}
% ==================================================================

>>>>

\begin{figure*}[t!]
\centering
\subfloat[\label{fig:northCrossing-1}{Crossing resolution vs no.~of vertices}]{
\centering
%======================
% Crossing Resolution
%======================
\begin{tikzpicture}
\begin{axis}[
axis x line*=bottom,xlabel style={yshift=0.2cm},ylabel style={yshift=-0.2cm},axis y line*=left,
legend style={at={(0.425,1.35)},anchor=north,,legend columns=2,draw=none},
width=0.475\textwidth,
cycle list name=mylist,
mark repeat={5},xlabel={Number of Vertices},ylabel={Crossing Resolution},
xtick={10,20,30,40,50,60,70,80,90,100},
xmin=10,
xmax=100,
ytick={10,20,30,40,50,60,70,80,90},
tick pos=left,ymajorgrids]
\addplot table [x=avg, y=cr rm, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=cr rm-1, col sep=semicolon] {crossingResolutionNorth.csv};
\addplot table [x=avg, y=cr fa, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=cr tr, col sep=semicolon] {crossingResolutionNorth.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}
\subfloat[\label{fig:northCrossing-2}{Aspect ratio vs no.~of vertices}]{
\centering
%===================================
% Crossing Resolution Aspect Ratio
%===================================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,legend style={at={(0.425,1.35)},
anchor=north,legend columns=2,draw=none},ylabel style={yshift=-0.2cm}, width=0.475\textwidth,cycle list name=mylist, mark repeat={5},xtick={0,10,20,30,40,50,60,70,80,90,100},ylabel={Aspect Ratio},
xlabel={Number of Vertices},
xmin=10,
xmax=100,
ytick={1,2,5,10,20,50,100,200,500,1000,2000},ymode=log, log ticks with fixed point,tick pos=left, ymajorgrids,]
\addplot table [x=avg, y=ar rm, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=ar rm-1, col sep=semicolon] {crossingResolutionNorth.csv};
\addplot table [x=avg, y=ar fa, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=ar tr, col sep=semicolon] {crossingResolutionNorth.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}

\subfloat[\label{fig:northCrossing-3}{No.~of crossings vs no.~of vertices}]{
\centering
%======================================
% Crossing Resolution Crossing Number
%======================================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},axis y line*=left,legend style={at={(0.425,1.35)},
anchor=north,legend columns=2,draw=none}, width=0.475\textwidth,cycle list name=mylist,  mark repeat={5},xtick={10,20,30,40,50,60,70,80,90,100},
xmin=10,
xmax=100,
ytick={0,250,500,750,1000,1250,1500,1750},ylabel={Number of Crossings},
xlabel={Number of Vertices},tick pos=left,ylabel style={yshift=-0.2cm}, ymajorgrids]
\addplot table [x=avg, y=cn rm, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=cn rm-1, col sep=semicolon] {crossingResolutionNorth.csv};
\addplot table [x=avg, y=cn fa, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=cn tr, col sep=semicolon] {crossingResolutionNorth.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}
\subfloat[\label{fig:northCrossing-4}{Iterations vs no.~of vertices}]{
\centering
%==================================
% Crossing Resolution Iterations
%==================================
\begin{tikzpicture}
\begin{axis}[axis x line*=bottom,xlabel style={yshift=0.2cm},ylabel style={yshift=-0.2cm},axis y line*=left,legend style={at={(0.425,1.35)},
anchor=north,legend columns=2,draw=none}, width=0.475\textwidth,cycle list name=mylist, mark repeat={5},xtick={10,20,30,40,50,60,70,80,90,100},
xmin=10,
xmax=100,
ytick={0,1000,2000,3000,4000,5000,6000,7000,8000},tick pos=left,
ylabel={Iterations},
xlabel={Number of Vertices},ymajorgrids]
\addplot table [x=avg, y=it rm, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=it rm-1, col sep=semicolon] {crossingResolutionNorth.csv};
\addplot table [x=avg, y=it fa, col sep=semicolon]{crossingResolutionNorth.csv};
\addplot table [x=avg, y=it tr, col sep=semicolon] {crossingResolutionNorth.csv};
\legend{{Unrestricted},{AR-restricted~~},{Huang et al.},{Argyriou et al.}}
\end{axis}
\end{tikzpicture}}
\caption{Experimental results for the crossing resolution on the AT\&T graphs.}
\label{fig:northCrossing}
\end{figure*}
